Tensor denoising and completion based on ordinal observations

### 2020-02-06 ICML/submit/ordinalTv2.tex
Add files via upload

\section{Introduction}
Multidimensional arrays, a.k.a.\ tensors, arise in a variety of applications including recommendation systems~\citep{adomavicius2011context}, social networks~\citep{nickel2011three}, genomics~\citep{hore2016tensor}, and neuroimaging~\citep{zhou2013tensor}. There is a growing need to develop dimension reduction methods for analyzing these noisy, high-dimensional datasets. The problem of tensor denoising -- which aims to recover a signal tensor from its noisy entries -- has gained increased attention in theory and applications~\citep{xia2019sup,zeng2019multiway}. A related problem, tensor completion, examines the minimum number of entries needed for a consistent recovery~\citep{ghadermarzy2018learning,ghadermarzy2019near}. Low-rankness is often imposed to the signal tensor, which efficiently reduces the intrinsic dimension in both problems.

A number of low-rank tensor estimation methods have been proposed~\citep{kolda2009tensor,acar2010scalable}, revitalizing classical methods such as CANDECOMP/PARAFAC (CP) decomposition~\citep{hitchcock1927expression} and Tucker decomposition~\citep{tucker1966some}. These tensor methods treat the entries as continuous-valued. In many cases, however, we encounter datasets of which the entries are qualitative. For example, the Netflix problem records the ratings of users to movies over weeks. Each data is a rating on a nominal scale \{{\it very like, like, neutral, dislike, very dislike}\}. Another example is in the signal processing, where the digits are frequently rounded or truncated so that only integer values are available. The qualitative observations take values in a limited set of categories, making the learning problem harder compared to the finer, continuous observations.

The analyses of ordinal tensors are complicated by several factors. Ordinal entries are categorical variables with an ordering among the categories; for example, {\it very like} $\prec$ {\it like} $\prec$ {\it neutral} $\prec \cdots$. A reasonable ordinal tensor model has two key properties. First, the model should be invariant under a reversal of categories, say, {\it very like} $\succ$ {\it like} $\succ$ {\it neutral} $\succ \cdots$, but not under arbitrary label permutations. Second, the parameter interpretations should be consistent under merging or splitting of contiguous categories. The classical continuous tensor model~\cite{kolda2009tensor, ghadermarzy2019near} fails in the first aspect, whereas the binary tensor model~\cite{ghadermarzy2018learning} lacks the second property. An appropriate model for ordinal tensors has yet to be studied.

@@

{\bf Our contribution.} This paper presents an efficient low-rank estimation method and theory for tensors with ordinal-valued entries. Our main contributions are summarized in Table~\ref{tab:compare}. We propose a cumulative link model for higher-order tensors, develop a rank-constrained M-estimator, and obtain theoretical accuracy guarantees. The mean squared error is established, and we show that the obtained rate is minimax optimal in high dimensions under the low-rank model. Our estimator enjoys a faster convergence rate $\tO(d^{-(K-1)/2})$ than $\tO(d^{-K})$ in~\citet{ghadermarzy2018learning}, which is a substantial improvement as the order $K$ increases. Furthermore, our proposal serves as an efficient completion algorithm that guarantees consistent recovery of an order-$K$ $(d,\ldots,d)$-dimensional low-rank tensor using only $\tO(Kd)$ noisy, quantized observations.

{\bf Related work.} Our work is related to, but clearly distinctive from, several lines of existing literature. Matrix completion from quantized samples was firstly introduced for binary observations~\cite{cai2013max,davenport2014,bhaskar20151} and then extended to ordinal observations~\cite{bhaskar2016probabilistic}. As we show in Section~\ref{sec:theory}, applying existing matrix methods to an ordinal tensor will result in a suboptimal estimator with a slower convergence rate. Therefore, a full exploitation of the tensor structure is necessary; this is the focus of the current paper.

Our work is also connected to non-Gaussian tensor decomposition. Existing work focuses exclusively on univariate observations such as binary- or continuous-valued entries~\citep{wang2018learning,hong2018generalized}. As we mentioned earlier, the ordinal observations adds considerable challenges to the model formulation. From statistical perspective, our proposed model generalizes the usual binary tensor model while preserving palindromic invariance~\cite{mccullagh1980regression} for ordinal observations. From algorithm perspective, our alternating optimization compares favorably to the approximate (non-convex) algorithm developed in the context of binary tensors~\cite{ghadermarzy2018learning}. We numerically compare the two approaches in Section~\ref{sec:experiment}.

@@

\subsection{Latent-variable interpretation}\label{sec:latent} 
We show that the ordinal tensor model~\eqref{eq:model} is equivalent to an $L$-level quantization model on $\tY=\entry{y_\omega}$:
\begin{align}\label{eq:quantization}
y_\omega&=
\begin{cases}
1,& \text{if $y^*_\omega\in(-\infty, b_1]$},\\
2,& \text{if $y^*_\omega\in(b_1, b_2]$},\\
\vdots  &\vdots\\
L,& \text{if $y^*_\omega\in(b_{L-1}, \infty)$},\\
\end{cases}
\end{align}
for all $\omega\in[d_1]\times \cdots \times [d_k]$. Here, $\tY^*=\entry{y^*_\omega}$ is a latent continuous-valued tensor following an additive noise model:
\begin{align}\label{eq:latent}
\KeepStyleUnderBrace{\tY^*}_{\text{latent continuous-valued tensor}}=\KeepStyleUnderBrace{\Theta}_{\text{signal tensor}}+\KeepStyleUnderBrace{\tE}_{\text{i.i.d.\ noise}},
\end{align}
where $\tE=\entry{\varepsilon_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a noise tensor with i.i.d.\ entries according to distribution $\mathbb{P}(\varepsilon)$. From the viewpoint of~\eqref{eq:latent}, the parameter tensor $\Theta$ can be interpreted as the latent signal tensor prior to contamination and quantization.

The equivalence between the latent-variable model~\eqref{eq:quantization} and the cumulative link model~\eqref{eq:model} is established if the link $f$ is chosen to be athe cumulative distribution function of noise $\varepsilon$, i.e., $f(\theta)=\mathbb{P}(\varepsilon\leq \theta)$. We describe two common choices of link $f$, or equivalently, the distribution of $\varepsilon$.

@@

(P1) Given the quantization process induced by $f$ and the cut-off points $\mb$, how accurate can we estimate the latent signal tensor $\Theta$ from the ordinal observation $\tY$?

@@

The parameter tensor of our interest satisfies two constraints. The first is that $\Theta$ is a low-rank tensor, with $r_k=\tO(1)$ for all $k\in[K]$. Equivalently, $\Theta$ admits the Tucker decomposition:
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor, $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns, and $\times_k$ denotes the tensor-by-matrix multiplication~\cite{kolda2009tensor}. The Tucker low-rankness is popularly imposed in tensor data analysis, and is shown to provide a reasonable tradeoff between model complexity and model flexibility. Note that, unlike matrices, there are various notations of tensor low-rankness, such as CP rank~\cite{hitchcock1927expression} and train rank~\cite{oseledets2011tensor}. Some notation of low-rankness may lead to mathematically ill-defined optimization; for example, the best low CP-rank tensor approximation may not exist~\cite{de2008tensor}. We choose Tucker representation for parsimony and easy interpretation.

@@

\section{Rank-constrained M-estimator}\label{sec:theory}
We present a general treatment to both problems mentioned above. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega=[d_1]\times \cdots \times [d_K]$ (for the tensor denoising) or a random subset induced from the sampling distribution $\Pi$ (for the tensor completion). Define $b_0=-\infty$, $b_L=\infty$, $f(-\infty)=0$ and $f(\infty)=1$. The log-likelihood associated with the observed entries is
\begin{align}\label{eq:objective}
\logl(\Theta, \mb)&=\sum_{\omega\in\Omega}\sum_{\ell\in[L]} \Big\{\mathds{1}_{\{y_\omega=\ell\}} \log \big[f(b_\ell-\theta_\omega)- \notag \\
&\quad \quad \quad \quad \quad \quad \quad  f(b_{\ell-1}-\theta_\omega)\big]\Big\}.
\end{align}
We propose a rank-constrained maximum likelihood estimator (a.k.a.\ M-estimator) for $\Theta$:
\begin{align}\label{eq:estimator}
\hat \Theta&=\argmax_{\Theta\in \tP}\logl(\Theta, \mb),\ \text{where}\notag \\
\tP&=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
 \end{align}
In practice, the cut-off points $\mb$ are unknown and should be jointly estimated with $\Theta$. For technical convenience, we assume in this section that the cut-off points $\mb$ are known. The adaptation of unknown $\mb$ is addressed in Section~\ref{sec:algorithm} and the Supplement. Similar treatment applies to $\mr$.

@@

\subsection{Estimation error for tensor denoising}\label{sec:denosing}
For the tensor denoising problem, we assume the full set of tensor entries are observed. We assess the estimation accuracy using the mean squared error (MSE):
\[
\text{MSE}(\hat \Theta, \trueT)={1\over \prod_k d_k}\FnormSize{}{\Theta-\trueT}^2.
\]
The next theorem establishes the upper bound for the proposed $\hat \Theta$ in~\eqref{eq:estimator}.

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\text{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_2  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1, c_2>0$ are two constants that depend only on $K$. 
\end{thm}
Theorem~\ref{thm:rate} establishes the statistical convergence for the estimator~\eqref{eq:estimator}. In fact, the proof of this theorem (see the Supplement) shows that the same statistical rate holds, not only for the global optimizer~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP: \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the objective value upon convergence is large enough. In Sections~\ref{sec:algorithm} and~\ref{sec:experiment}, we perform empirical studies to assess the algorithmic stability and its impact to statistical efficiency.

@@

Our bound also generalizes the previous results on ordinal matrices. The convergence rate for rank-constrained matrix estimation was $\tO(1/\sqrt{d})$~\citep{bhaskar2016probabilistic}, which fits into our special case when $K=2$. Furthermore, our results~\eqref{eq:rate} and~\eqref{eq:KLrate} reveal that the convergence becomes favorably as the order of data tensor increases. Intuitively, the sample size for tensor data analysis is the number of entries, $\prod_k d_k$, and the number of free parameters is roughly on the order of $\sum_{k}d_k$, assuming $r_{\max}=\tO(1)$. A higher tensor order implies higher effective sample size per parameter, and thus exhibits a faster convergence rate in high dimensions.

@@

Note that the recovery error depends on the distribution $\Pi$. In particular, tensor entries with higher sampling probabilities are equipped with better accuracy guarantee, compared to the ones with lower sampling probabilities.

@@

\begin{thm}~\label{thm:completion}
Assume the same set-up as in Theorem~\ref{thm:rate}. Suppose that we observe a subset of tensor entries $\{y_\omega\}_{\omega\in\Omega}$, where $\Omega$ is chosen at random with replacement according to a probability distribution $\Pi$. Let $\hat \Theta$ be the solution to~\eqref{eq:estimator}, and assume $r_{\max}=\tO(1)$. Then, with very high probability, 
\[
\PiFnormSize{}{\Theta-\hat \Theta}^2 \to 0,\quad \text{ as }\quad {|\Omega|\over \sum_k d_k} \ \to \infty. 
\]
\end{thm}
Theorem~\ref{thm:completion} shows that our estimator achieves consistent recovery using as few as $\tO(Kd)$ noisy, quantized observations from an order-$K$ $(d,\ldots,d)$-dimensional tensor. Note that $\tO(Kd)$ matches the degree of freedom for an order-$K$ tensor of fixed rank $\mr$, suggesting the optimality of our sample requirement. This sample complexity substantially improves over earlier result $\tO(d^{\lceil K/2\rceil})$ based on square matrixization~\cite{mu2014square}, or $\tO(d^{N/2})$ based on tensor nuclear-norm regularization~\cite{yuan2016tensor}. Existing methods that achieve $\tO(Kd)$ sample complexity require either a deterministic cross sampling design~\cite{zhang2019cross} or univariate measurements~\cite{ghadermarzy2018learning}. Our method extends the conclusion to multi-level measurements under a broader class of sampling schemes.

@@

\begin{figure}[H]
\centering
\vspace{-.5cm}
\includegraphics[width=8cm]{algorithm.pdf}\label{fig:stability}
\vspace{-.3cm}
\caption{Trajectory of objective function with various $d$ and $r$.}
\vspace{-.7cm}
\end{figure}

@@

The first experiment examines the performance under complete observations. We assess the empirical relationship between the MSE and various aspects of model complexity, such as dimension $d$, rank $r$, and signal level $\alpha=\mnormSize{}{\Theta}$. Figure~\ref{fig:finite}a plots the estimation error versus the tensor dimension $d$ for three different ranks $r\in\{3,5,8\}$. The decay in the error appears to behave on the order of $d^{-2}$, which is consistent with our theoretical results~\eqref{eq:rate}. It is seen that a higher rank leads to a larger error, as reflected by the upward shift of the curve as $r$ increases. Indeed, a higher rank implies a higher intrinsic dimension, thus increasing the difficulty of the estimation. Figure~\ref{fig:finite}b shows the estimation error versus the signal level under $d=20$. Interestingly, a larger estimation error is observed when the signal is either too small or too large. The non-monotonic behavior may seem surprising, but this is an intrinsic feature in the estimation with ordinal data. In view of the latent-variable interpretation (see Section~\ref{sec:latent}), estimation from ordinal observation can be interpreted as an inverse problem of quantization. Therefore, the estimation error diverges in the absence of noise $\tE$, because it is impossible to distinguish two different signal tensors, e.g., $\Theta_1=\ma_1\otimes \ma_2\otimes \ma_3$ and $\Theta_2=\text{sign}(\ma_1)\otimes \text{sign}(\ma_2)\otimes \text{sign}(\ma_3)$, from the quantized observations. This phenomenon~\cite{davenport2014,sur2019modern} is clearly contrary to the classical continuous-valued tensor problem.

The second experiment investigates the incomplete observations. We consider $L$-level tensors with $d=20$, $\alpha=10$ and choose a subset of tensor entries via uniform sampling. Figure~\ref{fig:finite}c shows the estimation error of $\hat \Theta$ versus the fraction of observation $\rho=|\Omega|/d^K$. As expected, the error reduces with increased $\rho$ or decreased $r$. Figure~\ref{fig:finite}d evaluates the impact of ordinal levels $L$ to estimation accuracy, under the setting $\rho=0.5$. An improved performance is observed as $L$ grows, especially from binary observations ($L=2$) to multi-level ordinal observations ($L\geq 3$). The result showcases the benefit of multi-level observations provide compared to binary observations.

@@

We apply each of the above methods to $L$-level ordinal tensors $\tY$ generated from model~\eqref{eq:model}. The {\bf Continuous-T} is applied directly to $\tY$ by treating the $L$ levels as continuous observations. The {\bf Ordinal-M} is applied to the matrix $\tY_{(1)}$ obtained via 1-mode unfolding. The {\bf 1bit-T} is applied to $\tY$ in two ways. The first approach ({\bf 1bit-sign-T}) follows from~\citet{ghadermarzy2018learning} that transforms $\tY$ to a binary tensor of equal size, by taking the entrywise sign of the mean-adjusted tensor, $\tY- |\Omega|^{-1}\sum_\omega y_\omega$. The second approach ({\bf 1bit-category-T}) transforms the order-3 ordinal tensor $\tY$ to an order-4 binary tensor $\tY^{\sharp}=\entry{y^{\sharp}_{ijkl}}$ via dummy variable encoding; i.e., $y^{\sharp}_{ijk\ell}=\mathds{1}_{\{y_{ijk}=\ell\}}$ for all $\ell\in[L-1]$. We evaluate the methods by their capabilities in predicting the most frequent label for each entry, i.e.,\ $y_\omega^{\text{mode}}=\arg\max_\ell\mathbb{P}(y_\omega=\ell)$. Two performance metrics are considered: mean absolute deviation, $\text{MAD}=d^{-K}\sum_\omega |y_\omega^{\text{mode}}-\hat y_\omega^{\text{mode}}|$, and misclassification rate, $\text{MCR}=d^{-K}\sum_\omega|y_\omega^{\text{mode}}-\text{round}(\hat y_\omega^{\text{mode}})|$, where $\text{round}(\cdot)$ denotes the nearest integer of the (possibly continuous-valued) prediction, e.g. returned by {\bf Continuous-T}. Note that MAD penalizes the large deviation more heavily than does MCR.

@@

\begin{figure}[t]
\includegraphics[width=8.3cm]{panel2.pdf}
\vspace{-.4cm}
\caption{Performance comparison in MSE (a-b) and MAD (c-d). (a, b) Prediction errors versus the number of ordinal levels $L$ when $\rho=0.8$. (c, d) Prediction errors versus sample complexity $\rho=|\Omega|/d^K$ when $L=5$. }\label{fig:compare}
\vspace{-.5cm}
\end{figure}

>>>>

\section{Introduction}
Multidimensional arrays, a.k.a.\ tensors, arise in a variety of applications including recommendation systems~\citep{adomavicius2011context}, social networks~\citep{nickel2011three}, genomics~\citep{hore2016tensor}, and neuroimaging~\citep{zhou2013tensor}. There is a growing need to develop general methods for analyzing these noisy, high-dimensional datasets that can handle two main problems. The problem of tensor denoising -- which aims to recover a signal tensor from its noisy entries -- has gained increased attention in theory and applications~\citep{xia2019sup,zeng2019multiway}. A related problem, tensor completion, examines the minimum number of entries needed for a consistent recovery~\citep{ghadermarzy2018learning,ghadermarzy2019near}. Low-rankness is often imposed to the signal tensor, which efficiently reduces the intrinsic dimension in both problems.

A number of low-rank tensor estimation methods have been proposed~\citep{kolda2009tensor,acar2010scalable}, revitalizing classical methods such as CANDECOMP/PARAFAC (CP) decomposition~\citep{hitchcock1927expression} and Tucker decomposition~\citep{tucker1966some}. These tensor methods treat the entries as continuous-valued. In many cases, however, we encounter datasets of which the entries are qualitative. For example, the Netflix problem records the ratings of users to movies over weeks. Each data is a rating on a nominal scale \{{\it very like, like, neutral, dislike, very dislike}\}. Another example is in the signal processing, where the digits are frequently rounded or truncated so that only integer values are available. The qualitative observations take values in a limited set of categories, making the learning problem harder compared to  continuous observations.

 Ordinal entries are categorical variables with an ordering among the categories; for example, {\it very like} $\prec$ {\it like} $\prec$ {\it neutral} $\prec \cdots$. The analyses of tensors with the ordinal entries are mainly complicated by two key properties needed for a reasonable model. First, the model should be invariant under a reversal of categories, say from the example, {\it very like} $\succ$ {\it like} $\succ$ {\it neutral} $\succ \cdots$, but not under arbitrary label permutations. Second, the parameter interpretations should be consistent under merging or splitting of contiguous categories. The classical continuous tensor model~\cite{kolda2009tensor, ghadermarzy2019near} fails in the first aspect, whereas the binary tensor model~\cite{ghadermarzy2018learning} lacks the second property. An appropriate model for ordinal tensors has not yet to be studied.

@@

{\bf Our contribution.} This paper presents an efficient low-rank estimation method and theory for tensors with ordinal-valued entries. Our main contributions are summarized in Table~\ref{tab:compare}. We propose a cumulative link model for higher-order tensors, develop a rank-constrained M-estimator, and obtain theoretical accuracy guarantees. The upper bound for the mean squared error is established, and we show that the obtained bound has minimax optimal rate in high dimensions under the low-rank model. Our estimator enjoys a faster convergence rate $\tO(d^{-(K-1)/2})$ than $\tO(d^{-K})$ in~\citet{ghadermarzy2018learning}, which is a substantial improvement as the order $K$ increases. Furthermore, our proposal serves as an efficient completion algorithm that guarantees consistent recovery of an order-$K$ $(d,\ldots,d)$-dimensional low-rank tensor using only $\tO(Kd)$ noisy, quantized observations.

{\bf Related work.} Our work is related to, but clearly distinctive from, several lines of existing literature. Matrix completion from quantized samples was firstly introduced for binary observations~\cite{cai2013max,davenport2014,bhaskar20151} and then extended to ordinal observations~\cite{bhaskar2016probabilistic}. As we show in Section~\ref{sec:theory}, applying existing matrix methods to an ordinal tensor results in a suboptimal estimator with a slower convergence rate. Therefore, a full exploitation of the tensor structure is necessary; this is the focus of the current paper.

Our work is also connected to non-Gaussian tensor decomposition. Existing work focuses exclusively on univariate observations such as binary- or continuous-valued entries~\citep{wang2018learning,hong2018generalized}. As we mentioned earlier, the ordinal observations add considerable challenges to the model formulation. We handle the problems from two perspectives. From statistical perspective, our proposed model generalizes the usual binary tensor model while preserving palindromic invariance~\cite{mccullagh1980regression} for ordinal observations. From algorithm perspective, our alternating optimization compares favorably to the approximate (non-convex) algorithm developed in the context of binary tensors~\cite{ghadermarzy2018learning}. We numerically compare the two approaches in Section~\ref{sec:experiment}.

@@

\subsection{Latent-variable interpretation}\label{sec:latent}
The ordinal tensor model~\eqref{eq:model} with certain types of link $f$ has the equivalent representation as an $L$-level quantization model on $\tY=\entry{y_\omega}$:
\begin{align}\label{eq:quantization}
y_\omega&=
\begin{cases}
1,& \text{if $y^*_\omega\in(-\infty, b_1]$},\\
2,& \text{if $y^*_\omega\in(b_1, b_2]$},\\
\vdots  &\vdots\\
L,& \text{if $y^*_\omega\in(b_{L-1}, \infty)$},\\
\end{cases}
\end{align}
for all $\omega\in[d_1]\times \cdots \times [d_k]$. Here, $\tY^*=\entry{y^*_\omega}$ is a latent continuous-valued tensor following an additive noise model:
\begin{align}\label{eq:latent}
\KeepStyleUnderBrace{\tY^*}_{\text{latent continuous-valued tensor}}=\KeepStyleUnderBrace{\Theta}_{\text{signal tensor}}+\KeepStyleUnderBrace{\tE}_{\text{i.i.d.\ noise}},
\end{align}
where $\tE=\entry{\varepsilon_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a noise tensor with i.i.d.\ entries according to distribution $\mathbb{P}(\varepsilon)$. From the viewpoint of~\eqref{eq:latent}, the parameter tensor $\Theta$ can be interpreted as the latent signal tensor prior to contamination and quantization.

The equivalence between the latent-variable model~\eqref{eq:quantization} and the cumulative link model~\eqref{eq:model} is established if the link $f$ is chosen to be the cumulative distribution function of noise $\varepsilon$, i.e., $f(\theta)=\mathbb{P}(\varepsilon\leq \theta)$. We describe two common choices of link $f$, or equivalently, the distribution of $\varepsilon$.

@@

(P1) Given the quantization process induced by $f$ and the cut-off points $\mb$, how accurately can we estimate the latent signal tensor $\Theta$ from the ordinal observation $\tY$?

@@

The parameter tensor of our interest satisfies two constraints. The first is that $\Theta$ is a low-rank tensor, with $r_k=\tO(1)$ for all $k\in[K]$. Equivalently, $\Theta$ admits the Tucker decomposition:
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor, $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns, and $\times_k$ denotes the tensor-by-matrix multiplication~\cite{kolda2009tensor}. The Tucker low-rankness is popularly imposed in tensor data analysis, and is shown to provide a reasonable tradeoff between model complexity and model flexibility. Note that, unlike matrices, there are various notations of tensor low-rankness, such as CP rank~\cite{hitchcock1927expression} and train rank~\cite{oseledets2011tensor}. Some notation of low-rankness may lead to mathematically ill-defined optimization; for example, the best low CP-rank tensor approximation may not exist~\cite{de2008tensor}. We choose Tucker representation for well definedness of optimization and easy interpretation.

@@

\section{Rank-constrained M-estimator}\label{sec:theory}
We present a general treatment to both problems mentioned above. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega=[d_1]\times \cdots \times [d_K]$ (for the tensor denoising) or a random subset induced from the sampling distribution $\Pi$ (for the tensor completion). Define $b_0=-\infty$, $b_L=\infty$, $f(-\infty)=0$ and $f(\infty)=1$. The log-likelihood associated with the observed entries is
\begin{align}\label{eq:objective}
\logl(\Theta, \mb)&=\sum_{\omega\in\Omega}\sum_{\ell\in[L]} \Big\{\mathds{1}_{\{y_\omega=\ell\}} \log \big[f(b_\ell-\theta_\omega)- \notag \\
&\quad \quad \quad \quad \quad \quad \quad  f(b_{\ell-1}-\theta_\omega)\big]\Big\}.
\end{align}
We propose a rank-constrained maximum likelihood estimator (a.k.a.\ M-estimator) for $\Theta$:
\begin{align}\label{eq:estimator}
\hat \Theta&=\argmax_{\Theta\in \tP}\logl(\Theta, \mb),\ \text{where}\notag \\
\tP&=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
 \end{align}
In practice, the cut-off points $\mb$ are unknown and should be jointly estimated with $\Theta$. For technical convenience, we assume in this section that the cut-off points $\mb$ are known. The adaptation of unknown $\mb$ is addressed in Section~\ref{sec:algorithm} and the Supplement.

@@

\subsection{Estimation error for tensor denoising}\label{sec:denosing}
For the tensor denoising problem, we assume that the full set of tensor entries are observed. We assess the estimation accuracy using the mean squared error (MSE):
\[
\text{MSE}(\hat \Theta, \trueT)={1\over \prod_k d_k}\FnormSize{}{\Theta-\trueT}^2.
\]
The next theorem establishes the upper bound for the MSE of the proposed $\hat \Theta$ in~\eqref{eq:estimator}.

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\text{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_2  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1, c_2>0$ are two constants that depend only on $K$.
\end{thm}
Theorem~\ref{thm:rate} establishes the statistical convergence for the estimator~\eqref{eq:estimator}. In fact, the proof of this theorem (see the Supplement) shows that the same statistical rate holds, not only for the global optimizer~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP: \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the convergent objective is large enough. In Section ~\ref{sec:algorithm}, we perform empirical studies to assess the algorithmic stability.

@@

Our bound also generalizes the previous results on ordinal matrices. The convergence rate for rank-constrained matrix estimation was $\tO(1/\sqrt{d})$~\citep{bhaskar2016probabilistic}, which fits into our special case when $K=2$. Furthermore, our results~\eqref{eq:rate} and~\eqref{eq:KLrate} reveal that the convergence becomes favorable as the order of data tensor increases. Intuitively, the sample size for tensor data analysis is the number of entries, $\prod_k d_k$, and the number of free parameters is roughly on the order of $\sum_{k}d_k$, assuming $r_{\max}=\tO(1)$. A higher tensor order implies higher effective sample size per parameter, and thus exhibits a faster convergence rate in high dimensions.

@@

Note that the recovery error depends on the distribution $\Pi$. In particular, tensor entries with higher sampling probabilities have more influence on the recovery accuracy, compared to the ones with lower sampling probabilities.

@@

\begin{thm}~\label{thm:completion}
Assume the same set-up as in Theorem~\ref{thm:rate}. Suppose that we observe a subset of tensor entries $\{y_\omega\}_{\omega\in\Omega}$, where $\Omega$ is chosen at random with replacement according to a probability distribution $\Pi$. Let $\hat \Theta$ be the solution to~\eqref{eq:estimator}, and assume $r_{\max}=\tO(1)$. Then, with very high probability,
\[
\PiFnormSize{}{\Theta-\hat \Theta}^2 \to 0,\quad \text{ as }\quad {|\Omega|\over \sum_k d_k} \ \to \infty.
\]
\end{thm}
Theorem~\ref{thm:completion} shows that our estimator achieves consistent recovery using as few as $\tO(Kd)$ {\color{red}($\tO(kd^{1+\epsilon})$?)} noisy, quantized observations from an order-$K$ $(d,\ldots,d)$-dimensional tensor. Note that $\tO(Kd)$ matches the degree of freedom for an order-$K$ tensor of fixed rank $\mr$, suggesting the optimality of our sample requirement. This sample complexity substantially improves over earlier result $\tO(d^{\lceil K/2\rceil})$ based on square matrixization~\cite{mu2014square}, or $\tO(d^{N/2})$ based on tensor nuclear-norm regularization~\cite{yuan2016tensor}. Existing methods that achieve $\tO(Kd)$ sample complexity require either a deterministic cross sampling design~\cite{zhang2019cross} or univariate measurements~\cite{ghadermarzy2018learning}. Our method extends the conclusions to multi-level measurements under a broader class of sampling schemes.

@@

\begin{figure}[H]
\centering
\vspace{-.4cm}
\includegraphics[width=8cm]{algorithm.pdf}\label{fig:stability}
\vspace{-.3cm}
\caption{Trajectory of objective function with various $d$ and $r$.}
\vspace{-.7cm}
\end{figure}

@@

The first experiment examines the performance under complete observations. We assess the empirical relationship between the MSE and various aspects of model complexity, such as dimension $d$, rank $r$, and signal level $\alpha=\mnormSize{}{\Theta}$. Figure~\ref{fig:finite}a plots the estimation error versus the tensor dimension $d$ for three different ranks $r\in\{3,5,8\}$. The decay in the error appears to behave on the order of $d^{-2}$, which is consistent with our theoretical results~\eqref{eq:rate}. It is seen that a higher rank leads to a larger error, as reflected by the upward shift of the curve as $r$ increases. Indeed, a higher rank implies the higher number of parameters to estimate, thus increasing the difficulty of the estimation. Figure~\ref{fig:finite}b shows the estimation error versus the signal level under $d=20$. Interestingly, a larger estimation error is observed when the signal is either too small or too large. The non-monotonic behavior may seem surprising, but this is an intrinsic feature in the estimation with ordinal data. In view of the latent-variable interpretation (see Section~\ref{sec:latent}), estimation from ordinal observation can be interpreted as an inverse problem of quantization. Therefore, the estimation error diverges in the absence of noise $\tE$, because it is impossible to distinguish two different signal tensors, e.g., $\Theta_1=\ma_1\otimes \ma_2\otimes \ma_3$ and $\Theta_2=\text{sign}(\ma_1)\otimes \text{sign}(\ma_2)\otimes \text{sign}(\ma_3)$, from the quantized observations. This phenomenon~\cite{davenport2014,sur2019modern} is clearly contrary to the classical continuous-valued tensor problem.

The second experiment investigates the incomplete observations. We consider $L$-level tensors with $d=20$, $\alpha=10$ and choose a subset of tensor entries via uniform sampling. Figure~\ref{fig:finite}c shows the estimation error of $\hat \Theta$ versus the fraction of observation $\rho=|\Omega|/d^K$. As expected, the error reduces with increased $\rho$ or decreased $r$. Figure~\ref{fig:finite}d evaluates the impact of ordinal levels $L$ to estimation accuracy, under the setting $\rho=0.5$. An improved performance is observed as $L$ grows, especially from binary observations ($L=2$) to multi-level ordinal observations ($L\geq 3$). The result showcases the benefit of multi-level observations compared to binary observations.

@@

We apply each of the above methods to $L$-level ordinal tensors $\tY$ generated from model~\eqref{eq:model}. The {\bf Continuous-T} is applied directly to $\tY$ by treating the $L$ levels as continuous observations. The {\bf Ordinal-M} is applied to the matrix $\tY_{(1)}$ obtained via 1-mode unfolding. The {\bf 1bit-T} is applied to $\tY$ in two ways. The first approach ({\bf 1bit-sign-T}) follows from~\citet{ghadermarzy2018learning} that transforms
$\tY$ to a binary tensor, by taking the entrywise sign of the mean-adjusted tensor, $\tY- |\Omega|^{-1}\sum_\omega y_\omega$. The second approach ({\bf 1bit-category-T}) transforms the order-3 ordinal tensor $\tY$ to an order-4 binary tensor $\tY^{\sharp}=\entry{y^{\sharp}_{ijkl}}$ via dummy variable encoding; i.e., $y^{\sharp}_{ijk\ell}=\mathds{1}_{\{y_{ijk}=\ell\}}$ for all $\ell\in[L-1]$.
We evaluate the methods by their capabilities in predicting the highest probability label for each entry, i.e.,\ $y_\omega^{\text{mode}}=\arg\max_\ell\mathbb{P}(y_\omega=\ell)$. Two performance metrics are considered: mean absolute deviation, $\text{MAD}=d^{-K}\sum_\omega |y_\omega^{\text{mode}}-\hat y_\omega^{\text{mode}}|$, and misclassification rate, $\text{MCR}=d^{-K}\sum_\omega\mathds{1}_{y_\omega^{\text{mode}}\neq\text{round}(\hat y_\omega^{\text{mode}})}$,
where $\text{round}(\cdot)$ denotes the nearest integer of the  prediction (possibly continuous-valued returned by {\bf Continuous-T}). Note that MAD penalizes the large deviation more heavily than MCR does.

@@

\begin{figure}[t]
\includegraphics[width=8.3cm]{panel2.pdf}
\vspace{-.4cm}
\caption{Performance comparison in MCR (a, b) and MAD (c, d). (b, d)   Prediction errors versus the number of ordinal levels $L$ when $\rho=0.8$.  (a, c) Prediction errors versus sample complexity $\rho=|\Omega|/d^K$ when $L=5$. }\label{fig:compare}
\vspace{-.5cm}
\end{figure}
